---
title: "Taxonomy plots"
author: "James Abbott"
date: "13/08/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '..')
library(dplyr)
library(tidyr)
library(ggplot2)
library(cowplot)

kraken_dir='parsed_outputs'
 
samples=c('Bulk_1','Bulk_2','Bulk_3','Desert_1','Desert_2','Desert_3',
          'North_1','North_2','North_3','Elite_1','Elite_2','Elite_3')
```

## Taxonomy plots

The parsed outputs produced from kraken using the kraken.ipynb notebooks are used to generate stacked bar-plots of taxonomic diversity in each sample at the phylum, class, order and family levels.

The plot_taxcomp function produces a single pdf file containing the four plots for i.e. bacteria, or fungi.

Required arguments: 
  * tax_data(named list): A list of data frames, one per rank, named with the rank. The dataframes contain three columns
    1: Sample name
    2: taxa (factor)
    3: count - reads assigned to taxa 

  * 'type': i.e. kraken - included in output filename to allow distinguishing between different analysis results

  * 'set': i.e fungi - included in output filename to allow distinguising between different subsets of data

Returns:
  * None
  
```{r per_sample_barplots}
plot_taxcomp<-function(tax_data,type,set) {
  levels=names(tax_data)
  plots=list() # list to store plots as they are generated for combining later
  
  #iterate through ranks and produce a plot for each... 
  for (level in levels) {
    level_dat<-tax_data[[level]]
    all_samples=data.frame(sample=character(0),taxa=character(0),count=numeric(0),pct=numeric(0))
    colnames(all_samples)<-colnames(level_dat)
    
    #determine %age compostion in each sample, storing results in all_samples dataframe
    for (sample in samples) {
      sample_df<-level_dat[level_dat$sample == sample,]
      sample_df<-head(sample_df[order(-sample_df$count),],20)
      sample_df$pct=sample_df$count/sum(sample_df$count)*100
      all_samples<-rbind(all_samples,sample_df)
    }
    
    # Selection of top-10 taxa by abundance for plotting:
    
    # Reshape to wide format, and move sample names to be row names
    wide<-pivot_wider(all_samples,names_from=taxa,values_from=c(count,pct))
    row.names(wide)<-wide$sample
    wide<-wide %>% select(- sample)
    # don't need the pct-ages for this stage
    wide<-wide %>% select(- starts_with('pct'))
    
    # Obtain totals for each column, sort on these and select the top 10, 
    # dropping rows with na's
    totals<-colSums(wide)
    totals<-sort(totals,decreasing=TRUE)
    totals<-totals[1:10]
    totals<-totals[!is.na(totals)]
    # Extract the names of the top 10 taxa and filter all_samples dataframe to only 
    # contain this subset
    top_taxa<-names(totals)
    top_taxa<-sapply(top_taxa,function(x) gsub('count_','',x))
    all_samples<-all_samples %>% filter(taxa %in% top_taxa)
    
    # split sample names to allow faceting replicates by sample
    all_samples<-all_samples %>% separate(col='sample',into=c('name','rep'),remove = FALSE)
    
    all_samples$sample<-factor(all_samples$sample, levels=samples) 
    all_samples$name<-factor(all_samples$name,levels=c('Bulk','Desert','North','Elite'))
    
    # and create the plot...
    plot_colours<-c('#989898','#E69F00','#55B3E9','#009D73','#EFE242','#0373B1','#D45E01','#CB79A6','#000000','#902CEC')
    
    plot<-ggplot(all_samples,aes(fill=taxa,y=pct,x=rep))+
        geom_bar(position="stack", stat="identity")+
        theme_bw()+
        scale_fill_manual(values=plot_colours) +
        theme(axis.text.x = element_text(angle = 45,hjust=1),
              strip.background = element_rect(colour='black', fill='white')) +
        ggtitle(paste0(toupper(substr(level, 1, 1)), substr(level, 2, nchar(level))))+
        xlab('Replicate')+ylab('Percentage')+facet_wrap(name ~ .,nrow=1)
      plots[[level]]=plot
  }
  
  #Combine plots into one and save...
  all_plots=plot_grid(plots[['phylum']],plots[['class']],plots[['order']],plots[['family']])
  ggsave(paste0(type,'_',set,'_taxplot','.pdf'),plot=all_plots,device='pdf',width=40,height=30,units='cm')
  
}
```

Plots are produced by reading the 'parsed_outputs' and combining all samples for a rank in a single dataframe. These are assigned to a named list which is passed to the plot_taxcomp() function defined above.

```{r read_parsedakraken}  
bacteria_kraken_data=list()
fungi_kraken_data=list()

for (type in c('bacteria','fungi')){
  for (taxa in c('phylum','class','order','family')) {
    tax_df<-data.frame('sample'=character(0), taxa=character(0),'count'=character(0))
    
    for (sample in samples) {
      
      fn=paste0(kraken_dir,'/',sample,'_',type,'_',taxa,'.txt')
      df<-read.csv(fn,sep="\t",header=1)
      
      colnames(df)<-tolower(colnames(df))
      df<-df %>% select(taxa,count)
      df['sample']=sample
      # reorder and rename to match expectations of plotting function...
      df<-df[,c(3,1,2)]
      colnames(df)<-c('sample','taxa','count')
      tax_df<-rbind(tax_df,df)
    }
    
    if (type=='bacteria') {
      bacteria_kraken_data[[taxa]]=tax_df
    } else {
      fungi_kraken_data[[taxa]]=tax_df
    }
  }
}

plot_taxcomp(fungi_kraken_data,'fungi','kraken')
plot_taxcomp(bacteria_kraken_data,'bacteria','kraken')
```


 